High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-23 doc/asplos2013/impl.tex
speeling

Events are created on demand and are {\em owned} by the node on which they
were created.  The creation of an event occurs with no inter-node communcation.
The space of event
handles is divided across the nodes at startup time, allowing each node to assign handles
to new events without the risk of conflicting with another node's assignments.  The static division
of event handles also permits any node to determine the owning node of an event
without any communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

>>>>

Events are created on demand and are {\em owned} by the node on which they
were created.  The creation of an event occurs with no inter-node communication.
The space of event
handles is divided across the nodes at start-up time, allowing each node to assign handles
to new events without the risk of conflicting with another node's assignments.  The static division
of event handles also permits any node to determine the owning node of an event
without any communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

### 2012-07-23 doc/asplos2013/interface.tex
speeling

Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitves that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  The
first object in the interface is a {\tt RegionMetaData} object (line 1).  Physical
regions created by the same region meta-data object belong to the same class.  Physical
regions are represented by {\tt RegionInstance} objects (line 17).  When creating a physical
region the client must specify the {\tt Memory} in which the physical region is going
to be allocated.  Memories are described in more detail in Section~\ref{subsec:machmodel}.
If there is not enough space for the physical region allocation a {\tt NO\_INST} will
be returned.  There is no automatic coherence of data between physical regions in the
same class.  It is the client's responsiblity to manage data coherence using copies
between physical regions (line 22-23).  Just like other operations copies can be
predicated on an event and return an event corresponding to completion.

>>>>

Between deferred locks and barriers (described in Section~\ref{subsec:events}) clients 
have access to the same set of synchronization primitives that they traditionally have
in both threading and bulk-synchronous interfaces.  However, these operations can now 
be composed with other asynchronous operations which is not possible in any other interface
of which we are aware.

@@

Listing~\ref{lst:regionapi} shows a subset of the interface for physical regions.  The
first object in the interface is a {\tt RegionMetaData} object (line 1).  Physical
regions created by the same region meta-data object belong to the same class.  Physical
regions are represented by {\tt RegionInstance} objects (line 17).  When creating a physical
region the client must specify the {\tt Memory} in which the physical region is going
to be allocated.  Memories are described in more detail in Section~\ref{subsec:machmodel}.
If there is not enough space for the physical region allocation a {\tt NO\_INST} will
be returned.  There is no automatic coherence of data between physical regions in the
same class.  It is the client's responsibility to manage data coherence using copies
between physical regions (line 22-23).  Just like other operations copies can be
predicated on an event and return an event corresponding to completion.

### 2012-07-23 doc/asplos2013/intro.tex
speeling

\item To support synchronization in our event system we introduce {\em deferred locks}, 
a novel synchronization primitive that operates in an asynchronous enviroment (Section~\ref{subsec:locks}).

\item We describe a physical region system that provides structural information about
data layout.  The structural information provided by physical regions enables an extensible
interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
further lantecy hiding (Section~\ref{subsec:phyreg}).

@@

\item We report on micro-benchmarks that are designed to stress
our implementation.  The results indiciate that our implementation approaches the performance of
the underlying hardware (Section~\ref{sec:micro}).

\item We demonstrate that our interface is powerful enough to support Legion \cite{Legion12}, a higher-level
programming system, and that real Legion applications programmed to our interface
are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demostrate
that the same applications written in a bulk-synchronous model incur performance penalites between 22-135\%.
\end{itemize}

>>>>

\item To support synchronization in our event system we introduce {\em deferred locks}, 
a novel synchronization primitive that operates in an asynchronous environment (Section~\ref{subsec:locks}).

\item We describe a physical region system that provides structural information about
data layout.  The structural information provided by physical regions enables an extensible
interface for conjoining operations on data (e.g. reductions) with asynchronous data movement for 
further latency hiding (Section~\ref{subsec:phyreg}).

@@

\item We report on micro-benchmarks that are designed to stress
our implementation.  The results indicate that our implementation approaches the performance of
the underlying hardware (Section~\ref{sec:micro}).

\item We demonstrate that our interface is powerful enough to support Legion \cite{Legion12}, a higher-level
programming system, and that real Legion applications programmed to our interface
are capable of achieving high-performance (Section~\ref{sec:apps}).  Furthermore, we demonstrate
that the same applications written in a bulk-synchronous model incur performance penalties between 22-135\%.
\end{itemize}

### 2012-07-23 doc/asplos2013/micro.tex
speeling

Our lock micro-benchmark is designed to measure the rate at which locks can be granted.
A parameterized number of locks are created per node and their handles are made available to
every node.  Each node then creates a parameterized number of {\em chains} of lock/unlock
request pairs, where each request pair is to a random lock, and is made dependent on the 
previous lock/unlock pair's completion.  The chains are made long enough that startup
and cleanup times are negligible.  The total number of chains across all nodes gives the total
number of lock requests that can exist in the system at any given time.  All chains are
started at the same time (via a dependence on a single user event), and the total time to
process all the chains is divided into the total number of lock requests to yield and average
lock grant rate.

@@

The dense experiment demonstrates that reduction fold regions perform 
best and scale well with the number of nodes, achieving over a billion reductions
per second in the 16 node case.  List regions also scale well, but perform about an
order of mangitude worse than folding regions in the dense case.  The use of a separate active message for
each reduction operation is another two orders of magnitude worse, demonstrating that data must be
transferred in larger blocks to be efficient on modern interconnect hardware.  The localized instance approach
works well for a single node, but its inherently serial nature doesn't benefit at all from increasing node counts.  Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
on a single node (where the reads and writes are local).

>>>>

Our lock micro-benchmark is designed to measure the rate at which locks can be granted.
A parameterized number of locks are created per node and their handles are made available to
every node.  Each node then creates a parameterized number of {\em chains} of lock/unlock
request pairs, where each request pair is to a random lock, and is made dependent on the 
previous lock/unlock pair's completion.  The chains are made long enough that start-up
and cleanup times are negligible.  The total number of chains across all nodes gives the total
number of lock requests that can exist in the system at any given time.  All chains are
started at the same time (via a dependence on a single user event), and the total time to
process all the chains is divided into the total number of lock requests to yield and average
lock grant rate.

@@

The dense experiment demonstrates that reduction fold regions perform 
best and scale well with the number of nodes, achieving over a billion reductions
per second in the 16 node case.  List regions also scale well, but perform about an
order of magnitude worse than folding regions in the dense case.  The use of a separate active message for
each reduction operation is another two orders of magnitude worse, demonstrating that data must be
transferred in larger blocks to be efficient on modern interconnect hardware.  The localized instance approach
works well for a single node, but its inherently serial nature doesn't benefit at all from increasing node counts.  Finally, the only time the latency of performing individual RMA reads and writes isn't disastrous is
on a single node (where the reads and writes are local).

### 2012-07-23 doc/asplos2013/related.tex
speeling

Both POSIX threads and OpenMP\cite{OPENMP98} are used for intra-node parallel
programming in large machines, but neither support any asynchronous
operations.  CUDA\cite{CUDA} and OpenCL\cite{Khronos:OpenCL} support
the compostition of asynchronous kernel launches and asynchronous copies
between a host node and a single GPU.  However, the only synchronization
options available in both interfaces are blocking operations on either
a stream or a work queue.  OpenCL has events similar to our interface for
expressing ordering, but has no synchronization primitive for expressing
more relaxed properties such as atomicity that can be described by a
deferred lock.  OpenCL events are not valid globally, but can only be
used with a single GPU context.

>>>>

Both POSIX threads and OpenMP\cite{OPENMP98} are used for intra-node parallel
programming in large machines, but neither support any asynchronous
operations.  CUDA\cite{CUDA} and OpenCL\cite{Khronos:OpenCL} support
the composition of asynchronous kernel launches and asynchronous copies
between a host node and a single GPU.  However, the only synchronization
options available in both interfaces are blocking operations on either
a stream or a work queue.  OpenCL has events similar to our interface for
expressing ordering, but has no synchronization primitive for expressing
more relaxed properties such as atomicity that can be described by a
deferred lock.  OpenCL events are not valid globally, but can only be
used with a single GPU context.

### 2012-07-23 doc/asplos2013/results.tex
Removing a latex warning
speeling

The fully asynchronous design of the interface proposed in
this paper is crucial to Legion's SOOP implementation.  Without a fully
asynchronous interface, the SOOP would have to periodically invoke
blocking operations that would cause processors to stall and severly
limit Legion's ability to run ahead and keep the machine fully
occupied.  With a fully asynchronous interface the SOOP can hide latencies by 
running
far ahead of the actual execution, limited only by the physical resources
available in the machine, exactly like a hardware out-of-order processor.

@@

{\em Circuit} is a simulation of an integrated circuit that is described as an
irregular graph of components connected by wires.  The graph is partitioned
and distributed across the machine.  The computation for each time step consists
of three phases that calculate currents on the wires, distribute charge between
nodes, and update the voltages of all nodes.  In the distribute charge
phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
are used to allow each partition to perform reductions in parallel and locally in GPU
framebuffer memeory before merging all the reductions to a global instance
residing in GASNet memory.  Without reduction instances this phase
couldn't be performed using GPUs.

{\em Fluid} is an incompressible fluid flow simulation from the PARSEC
benchmark suite\cite{bienia11benchmarking}.  The original implementation can 
only be run on a single shared memory machine, but when written in Legion, Fluid can
be run on large distributed machines as well.  The Fluid simulation
models particles that flow through cells.  Each time step requires multiple phases
that update different proprerties of the particles contingent upon neighboring
particles in the same cell.  The space of cells is partitioned and neighboring
cells in different partitions must exchange data between phases.  Legion ensures
that these exchanges are done point-to-point by chaining copies and tasks
using events rather than employing a global bulk-synchronous approach.

@@

We instrumented the heterogenous implementation of our interface to 
capture event usage information.  Figure~\ref{fig:eventlife} shows
a timeline of the execution of the Fluid application on 16 nodes.
The dynamic events line measures the total number of event creations.
A large number of events are created - over 260,000 in less than 15
seconds - and allocating separate storage for every event would clearly
be difficult for long-running applications.

Dynamic events are considered {\em live events} until their last operation 
(i.e. wait, trigger) has been performed.  The live events line
measures the number of events that are live at any given time.  After
the last operation on an event, a reference counting implementation would
be able to recover the storage associated with the event.  In this example 
a reference counting algorithm would reduce the storage needed for dynanmic events
by over 10X, but could only do so at the cost of
computation and communication for performing the reference counting.

As discussed in Section~\ref{subsec:eventimpl}, our implementation
requires storage that grows with the maximum number of {\em untriggered events}, a number
that is 10X smaller than even the maximal live event count.  The actual storage
requirements of our implementation are shown by the {\em physical events} line,
which is slightly larger than the peak number of untriggered events. 
The reason for this difference is because nodes create events locally if they have
no free generational events, even if there are unused generational events on remote nodes.
Despite this inefficiency, our implementation uses 5X less storage
than a reference counting implementation and avoids any overhead related to 
reference counting.  These savings would likely be even more dramatic for longer 
runs of the application, as the number of live events is steadily growing as the
application runs, while the peak number of generational events needed appears to 
occur during the startup of the application.

@@

\begin{figure}[h!]
\centering
\subfigure[Circuit Application]{
\label{fig:cktbulk}
\includegraphics[scale=0.33]{figs/circuit_bulk_sync.pdf}
}

>>>>

The fully asynchronous design of the interface proposed in
this paper is crucial to Legion's SOOP implementation.  Without a fully
asynchronous interface, the SOOP would have to periodically invoke
blocking operations that would cause processors to stall and severely
limit Legion's ability to run ahead and keep the machine fully
occupied.  With a fully asynchronous interface the SOOP can hide latencies by 
running
far ahead of the actual execution, limited only by the physical resources
available in the machine, exactly like a hardware out-of-order processor.

@@

{\em Circuit} is a simulation of an integrated circuit that is described as an
irregular graph of components connected by wires.  The graph is partitioned
and distributed across the machine.  The computation for each time step consists
of three phases that calculate currents on the wires, distribute charge between
nodes, and update the voltages of all nodes.  In the distribute charge
phase, a single node may be updated by wires from multiple partitions.  Reduction instances 
are used to allow each partition to perform reductions in parallel and locally in GPU
framebuffer memory before merging all the reductions to a global instance
residing in GASNet memory.  Without reduction instances this phase
couldn't be performed using GPUs.

{\em Fluid} is an incompressible fluid flow simulation from the PARSEC
benchmark suite\cite{bienia11benchmarking}.  The original implementation can 
only be run on a single shared memory machine, but when written in Legion, Fluid can
be run on large distributed machines as well.  The Fluid simulation
models particles that flow through cells.  Each time step requires multiple phases
that update different properties of the particles contingent upon neighboring
particles in the same cell.  The space of cells is partitioned and neighboring
cells in different partitions must exchange data between phases.  Legion ensures
that these exchanges are done point-to-point by chaining copies and tasks
using events rather than employing a global bulk-synchronous approach.

@@

We instrumented the heterogeneous implementation of our interface to 
capture event usage information.  Figure~\ref{fig:eventlife} shows
a timeline of the execution of the Fluid application on 16 nodes.
The dynamic events line measures the total number of event creations.
A large number of events are created - over 260,000 in less than 15
seconds - and allocating separate storage for every event would clearly
be difficult for long-running applications.

Dynamic events are considered {\em live events} until their last operation 
(i.e. wait, trigger) has been performed.  The live events line
measures the number of events that are live at any given time.  After
the last operation on an event, a reference counting implementation would
be able to recover the storage associated with the event.  In this example 
a reference counting algorithm would reduce the storage needed for dynamic events
by over 10X, but could only do so at the cost of
computation and communication for performing the reference counting.

As discussed in Section~\ref{subsec:eventimpl}, our implementation
requires storage that grows with the maximum number of {\em untriggered events}, a number
that is 10X smaller than even the maximal live event count.  The actual storage
requirements of our implementation are shown by the {\em physical events} line,
which is slightly larger than the peak number of untriggered events. 
The reason for this difference is because nodes create events locally if they have
no free generational events, even if there are unused generational events on remote nodes.
Despite this inefficiency, our implementation uses 5X less storage
than a reference counting implementation and avoids any overhead related to 
reference counting.  These savings would likely be even more dramatic for longer 
runs of the application, as the number of live events is steadily growing as the
application runs, while the peak number of generational events needed appears to 
occur during the start-up of the application.

@@

\begin{figure}[!ht]
\centering
\subfigure[Circuit Application]{
\label{fig:cktbulk}
\includegraphics[scale=0.33]{figs/circuit_bulk_sync.pdf}
}
